---
title: "processing_basic"
author: "ozzie"
date: "2022-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r libs_and_data, include=FALSE}
library(tidyverse)
library(ggpubr)
library(phyloseq)
library(curatedMetagenomicData)
library(DT)
library(mia)

library(phangorn)

library(ape)

library(ggtree)

## read in your tree
basic_tree <- ape::read.tree("C:/Users/hammera/Documents/R_projects/Microbiopics/data/asv_prunedguides_rooted.nwk")

# count data table for the asvs in the tree
fresh_asv_table <- read.table("C:/Users/hammera/Documents/R_projects/Microbiopics/data/asv_table_new.tab", row.names=1, header=TRUE)

basic_asv_table <- fresh_asv_table/rowSums(fresh_asv_table)
```

```{r custom_functions, include=FALSE}
## needs to be ultrametric
make_ultrametric_tree <- function(new_tree){
  if(is.ultrametric(new_tree)){
    return(new_tree)
  }
  else{
    return(ape::chronos(new_tree, quiet=T, lambda=0))
  }
}

# need to define the minimum number of sub-clustered groups that are still informative
# basically this script just cuts the tree at a bunch of different points, and you can
# use the output to determine the point at which you think there isn't a lot of information being given. You may consider plotting this then just choosing an elbow point
determine_tree_chopping_point <- function(your_hclust){
  max_branch_len <- max(your_hclust$height)
  tree_groups <- sapply(1:1000, function(x) length(unique(cutree(your_hclust, h=x*0.0001*max_branch_len))))
  return(tree_groups)
}

# this function iterates over the hclust object and cuts the tree into small
# bits, returning the group membership number for each asv at each cut
# at finally returning a table of ASV membership at each cut over the whole tree
axe_the_tree <- function(tree_of_choice, number_cuts, max_stump_position){
  if(missing(max_stump_position)){
    max_stump_position=1
  }
  if(missing(number_cuts)){
    number_cuts=200
  }
  cut_step_length = (max_stump_position/number_cuts)
  membership_at_cuts <- sapply(1:number_cuts, function(x) cutree(tree_of_choice, h=x*cut_step_length))
  return(membership_at_cuts)
}

# this function takes an asv table, the output of a call to axe_the_tree
# and iterates over one sample and one column to generate the grouped output counts
# this should be called using sapply or some other iterative option to produce the 
# microbe image for all the data
make_tree_image <- function(the_asv_table, the_grouped_output, group_level, asv_sample_row){
  grouped_tree_slice <- data.frame(ASVs = colnames(the_asv_table),
                         asv_counts = as.numeric(as.vector(the_asv_table[asv_sample_row,]))) %>%
  inner_join(data.frame(ASVs = rownames(the_grouped_output),
                        groups = the_grouped_output[,group_level]), by="ASVs") %>%
  group_by(groups) %>%
  mutate(group_count = sum(asv_counts)) %>%
  ungroup()
  return(grouped_tree_slice$group_count)
}


# this function will cut the tree at pre-determined points
# this was optimized for cutting at points where the change in ASV number was greatest
cut_at_specific_points <- function(tree_of_choice, number_cuts, cut_position){
  membership_at_cuts <- sapply(1:number_cuts, function(x) cutree(tree_of_choice, h=cut_position[x]))
  return(membership_at_cuts)
}
```

```{r tree_cutters}
ultra_tree <- make_ultrametric_tree(basic_tree)

tclust <- as.hclust(ultra_tree)

tree_group_points <- determine_tree_chopping_point(tclust)

tree_chopping_df <- data.frame(number_of_groups = tree_group_points, operator=seq(1,1000)*0.002)

tree_chopping_df %>%
  ggplot(aes(x=operator, y=number_of_groups)) +
  geom_smooth(se=F)+
  theme_linedraw() +
  xlab("Position on Tree") +
  ylab("Number of Groups at Cut")

test_obj <- data.frame(sapply(1:100, function(x) cutree(tclust, h=x*(0.5/100))))

first_image <- data.frame(sapply(1:ncol(test_obj), function(x) make_tree_image(basic_asv_table, test_obj, x, 1)))
rownames(first_image) <- colnames(basic_asv_table)

first_image_ordered <- first_image[match(rownames(first_image), ultra_tree$tip.label),]

second_image <- data.frame(sapply(1:ncol(test_obj), function(x) make_tree_image(basic_asv_table, test_obj, x, 2)))
rownames(second_image) <- colnames(basic_asv_table)

second_image_ordered <- second_image[match(rownames(second_image), ultra_tree$tip.label),]

burning_tree_plot_s1 <- first_image_ordered %>%
  mutate(ASV = rownames(first_image_ordered)) %>%
  pivot_longer(!ASV, names_to="tree_cut", values_to="asv_group") %>%
  ggplot(aes(x=factor(tree_cut, level=c(colnames(first_image_ordered))), y=factor(ASV, level=c(basic_tree$tip.label)), fill=asv_group)) +
  geom_tile() +
  scale_fill_gradient(high="gold", low="purple") +
  theme(axis.text.x=element_blank(),
        axis.line.x = element_blank()) +
  xlab(label=NULL)


burning_tree_plot_s2 <- second_image_ordered %>%
  mutate(ASV = rownames(second_image_ordered)) %>%
  pivot_longer(!ASV, names_to="tree_cut", values_to="asv_group") %>%
  ggplot(aes(x=factor(tree_cut, level=c(colnames(second_image_ordered))), y=factor(ASV, level=c(basic_tree$tip.label)), fill=asv_group)) +
  geom_tile() +
  scale_fill_gradient(high="gold", low="purple") +
  theme(axis.text.x=element_blank(),
        axis.line.x = element_blank()) +
  xlab(label=NULL)

comb_plot <- ggarrange(burning_tree_plot_s1, burning_tree_plot_s2)

```

```{r real_data, include=FALSE}
sample_metadata <- curatedMetagenomicData::sampleMetadata
some_object <- curatedMetagenomicData("AsnicarF_20.+.relative_abundance", dryrun = TRUE, counts = TRUE, rownames = "short")
tiny_obj <- some_object$`2021-03-31.AsnicarF_2021.relative_abundance`
metagenomic_data <- some_object
row_tree <- assays(tiny_obj)
asv_count_table <- (row_tree$relative_abundance)
microbe_names <- gsub(" ", "_", rownames(asv_count_table))
micro_names <- gsub("\\.", "", microbe_names)



full_tree <- rowTree(tiny_obj)
full_tree_names <- full_tree$tip.label

modified_tree_names <- data.frame(t(data.frame(str_split(full_tree_names, "s__", n=2))))
colnames(modified_tree_names) <- c("more_tax", "species") 
rownames(modified_tree_names) <- NULL

tree_names <- modified_tree_names %>%
  mutate(fixed_names = gsub("\\:", "_", species))

count_table_names <- data.frame(microbe_names = micro_names) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\.", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, " ", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\:", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "-", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\[", "")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\]", ""))

count_table_names$microbe_names[!(count_table_names$microbe_names %in% tree_names$fixed_names)]


asv_count_table[rownames(asv_count_table)%in%tree_names$fixed_names]



# some add
```

```{r}
curated_adult_fecal_samples <- sampleMetadata %>%
  filter(age >= 18) %>%
    filter(body_site == "stool") %>%
    select(where(~ !all(is.na(.x))))
#    returnSamples("relative_abundance", rownames = "short")
    
study_samples <- sampleMetadata %>%
  filter(age >= 18) %>%
    filter(body_site == "stool") %>%
    select(where(~ !all(is.na(.x)))) %>%
  group_by(study_name) %>%
  summarize(samples_in_study = n())
    
grouped_conditions <- curated_adult_fecal_samples %>%
  group_by(study_condition) %>%
  summarise(sample_total = n())

gene_family_data <- sampleMetadata %>%
  filter(age >= 18) %>%
    filter(body_site == "stool") %>%
    select(where(~ !all(is.na(.x)))) %>%
  filter(study_name=="HanniganGD_2017")
#    returnSamples("gene_families", rownames = "short")


# the line below reads in data where curatedMetagenomicData was collected for all included adults that have some identifying metadata available
adults_rel_abund <- readRDS("C:/Users/hammera/Documents/R_projects/Microbiopics/data/adults_with_known_data_big_phyloseq.rds")

cmd_count_table <- data.frame(otu_table(adults_rel_abund))
count_sums <- rowSums(cmd_count_table)

#sample_gene_families_Hannigan2017 <- #readRDS("C:/Users/hammera/Documents/R_projects/Microbiopics/data/sample_gene_families_Hannigan2017.rds")
#gene_family_counts <- data.frame(otu_table(sample_gene_families_Hannigan2017))
#gene_family_names <- rownames(gene_family_counts)
#gene_family_sums <- rowSums(gene_family_counts)

sample_md <- (data.frame(sample_data(adults_rel_abund)))

grouped_study_condition <- sample_md %>%
  group_by(study_condition) %>%
  summarize(study_condition_total = n())
```

```{r checking_phylo_tree_edges}
count_tab_names <- data.frame(microbe_names = rownames(cmd_count_table)) %>%
  mutate(microbe_names = str_replace_all(microbe_names, " sp.", " sp")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\.", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, " ", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\:", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "-", "_")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\[", "")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "\\]", "")) %>%
  mutate(microbe_names = str_replace_all(microbe_names, "__", "_")) %>%
  mutate(original_names = rownames(cmd_count_table))

metaphlan_tree <- ape::read.tree("C:/Users/hammera/Documents/R_projects/Microbiopics/data/mpa_v30_CHOCOPhlAn_201901_species_tree.nwk")

all_metaphlan_tree_names <- metaphlan_tree$tip.label

modified_metaphlan_tree_names <- data.frame(t(data.frame(str_split(all_metaphlan_tree_names, "s__", n=2))))
colnames(modified_metaphlan_tree_names) <- c("more_tax", "species") 
rownames(modified_metaphlan_tree_names) <- NULL

metaphlan_tree_names <- modified_metaphlan_tree_names %>%
  mutate(microbe_names = gsub("\\:", "_", species)) %>%
  mutate(tree_names = all_metaphlan_tree_names)

# need to change the name for a couple of bugs in the metaphlan tree because the name doesn't match up to the count table and the taxonomic identifier has changed
metaphlan_tree_names$microbe_names[metaphlan_tree_names$microbe_names=="Bacteroides_vulgatus"] <- "Phocaeicola_vulgatus"
metaphlan_tree_names$microbe_names[metaphlan_tree_names$microbe_names=="Bacteroides_dorei"] <- "Phocaeicola_dorei"
metaphlan_tree_names$microbe_names[metaphlan_tree_names$microbe_names=="Bacteroides_plebeius"] <- "Phocaeicola_plebeius"
metaphlan_tree_names$microbe_names[metaphlan_tree_names$microbe_names=="Eubacterium_eligens"] <- "Lachnospira_eligens"
metaphlan_tree_names$microbe_names[metaphlan_tree_names$microbe_names=="Eubacterium_hallii"] <- "Anaerobutyricum_hallii"

# MAJOR NOTE: Not all of the colsums will be 100 (percent) because there are a small number of taxa, most of which are extremely unabundant, which do not have a mapping in the phylogenetic tree
# this issue has been noted on the biobakery github page for metaphlan4 and I think it's being worked out

# identify which names from the metphlan tree match to the count table and which ones don't
no_match_names <- count_tab_names[!(count_tab_names$microbe_names %in% metaphlan_tree_names$microbe_names),]
matched_names <- count_tab_names[(count_tab_names$microbe_names %in% metaphlan_tree_names$microbe_names),] %>%
  left_join(metaphlan_tree_names)


# filter to get just the rownames that are not in the table
no_match_count_table <- cmd_count_table %>%
  filter(rownames(cmd_count_table) %in% no_match_names$original_names)

microbes_not_matching <- sort(rowSums(no_match_count_table))
hist(rowSums(no_match_count_table), breaks=40)


split_names <- data.frame(strsplit(metaphlan_tree$tip.label, "s__"))
all_metaphlan_taxa <- data.frame(microbes = as.vector(split_names[2,])) %>%
  pivot_longer(everything(), names_to=NULL, values_to="microbes")

(no_match_names) %>% head(30)

metaphlan_tree_names$microbe_names[grepl("massiliensis", metaphlan_tree_names$microbe_names, ignore.case=TRUE)]

count_tab_names$microbe_names[(grepl("massiliensis", count_tab_names$microbe_names))]
```
```{r taking_microbiopics, include=FALSE}
# this is a df of the original count table for these data
cmd_count_table

# this is the og tree
metaphlan_tree
mtree <- metaphlan_tree

# first, select the tree names that are matched between the count table and the tree, then
# rename the tree labels so that they match those that are in the count table
# Finally, rename all the stuff

matched_tree <- ape::keep.tip(mtree, matched_names$tree_names)

matched_subsetted_count_table <- cmd_count_table %>%
  filter(rownames(cmd_count_table) %in% matched_names$original_names)

matched_colsums <- colSums(matched_count_table)

# organize the names so that they are in the same order as the tree tips, then rename the tree tips
tree_matched_df <- data.frame(tree_names = matched_tree$tip.label) %>%
  left_join(matched_names)

matched_tree$tip.label <- tree_matched_df$microbe_names

# this just reorganizes the names of the count table so that they match the data.frame
matched_count_df <- data.frame(original_names = rownames(matched_subsetted_count_table)) %>%
  left_join(matched_names)

matched_count_table <- matched_subsetted_count_table
rownames(matched_count_table) <- matched_count_df$microbe_names # uses same names for tree and count table
```

```{r build_tree_images}
sum_microbe_abundances <- data.frame("taxa_sums" = rowSums(matched_count_table)) %>%
rownames_to_column(var="microbes") %>%
  arrange(desc(taxa_sums)) %>%
  top_n(500)

filtered_tree <- ape::keep.tip(matched_tree, sum_microbe_abundances$microbes)

flipped_count_table <- matched_count_table %>%
  rownames_to_column(var="microbes") %>%
  pivot_longer(!microbes,
               values_to="counts",
               names_to = "Samples") %>%
  pivot_wider(names_from = microbes,
              values_from = counts) %>%
  column_to_rownames(var="Samples") %>%
  dplyr::select(sum_microbe_abundances$microbes)



filtered_ultra_tree <- make_ultrametric_tree(filtered_tree)

filter_clust <- as.hclust(filtered_ultra_tree)

tree_group_points <- determine_tree_chopping_point(filter_clust)

tree_chopping_df <- data.frame(number_of_groups = tree_group_points, operator=seq(1,1000)*0.002) %>%
  mutate(previous_step_taxa_number = c(500,(number_of_groups[1:999]))) %>%
  mutate(change_in_taxa = previous_step_taxa_number-number_of_groups) %>%
  arrange(desc(change_in_taxa)) %>%
  top_n(80) %>%
  arrange(operator)


tree_chopping_df %>%
  ggplot(aes(x=operator, y=number_of_groups)) +
  geom_smooth(se=F)+
  theme_linedraw() +
  xlab("Position on Tree") +
  ylab("Number of Groups at Cut")


test_tree_cut <- data.frame(sapply(1:length(tree_chopping_df$operator), function(x) cutree(filter_clust, h=tree_chopping_df$operator[x])))


first_image <- data.frame(sapply(1:ncol(test_tree_cut), function(x) make_tree_image(flipped_count_table, test_tree_cut, x, 1)))
rownames(first_image) <- colnames(flipped_count_table)

first_image_ordered <- first_image[match(rownames(first_image), filtered_tree$tip.label),]


second_image <- data.frame(sapply(1:ncol(test_tree_cut), function(x) make_tree_image(flipped_count_table, test_tree_cut, x, 2)))
rownames(second_image) <- colnames(flipped_count_table)

second_image_ordered <- second_image[match(rownames(second_image), filtered_tree$tip.label),]

burning_tree_plot_s1 <- first_image_ordered %>%
  mutate(ASV = rownames(first_image_ordered)) %>%
  pivot_longer(!ASV, names_to="tree_cut", values_to="asv_group") %>%
  ggplot(aes(x=factor(tree_cut, level=c(colnames(first_image_ordered))), y=factor(ASV, level=c(filtered_tree$tip.label)), fill=asv_group)) +
  geom_tile() +
  scale_fill_gradient(high="darkblue", low="white") +
  theme(axis.text.x=element_blank(),
        axis.line.x = element_blank()) +
  xlab(label=NULL)

burning_tree_plot_s2 <- second_image_ordered %>%
  mutate(ASV = rownames(second_image_ordered)) %>%
  pivot_longer(!ASV, names_to="tree_cut", values_to="asv_group") %>%
  ggplot(aes(x=factor(tree_cut, level=c(colnames(second_image_ordered))), y=factor(ASV, level=c(filtered_tree$tip.label)), fill=asv_group)) +
  geom_tile() +
  scale_fill_gradient(high="darkblue", low="white") +
  theme(axis.text.x=element_blank(),
        axis.line.x = element_blank()) +
  xlab(label=NULL)

comb_plot <- ggarrange(burning_tree_plot_s1, burning_tree_plot_s2)

for(sample_num in 1:nrow(flipped_count_table)){
  first_image <- data.frame(sapply(1:ncol(test_tree_cut), function(x) make_tree_image(flipped_count_table, test_tree_cut, x, sample_num)))
  rownames(first_image) <- colnames(flipped_count_table)
  first_image_ordered <- first_image[match(rownames(first_image), filtered_tree$tip.label),]
  new_data_object_name <- paste0("C:/Users/hammera/Documents/R_projects/Microbiopics/results/sample_pics/pic_dim_500_by_81/image_", sample_num, ".rds")
  saveRDS(first_image_ordered, new_data_object_name)
}

make_image_list <- function(your_tree, your_count_table, tree_cutting_plan, sample_index){
  an_image <- data.frame(sapply(1:ncol(tree_cutting_plan), function(x) make_tree_image(your_count_table, tree_cutting_plan, x, sample_index)))
  rownames(an_image) <- colnames(your_count_table)
  an_ordered_image <- an_image[match(rownames(an_image), your_tree$tip.label),]
  return(an_ordered_image)
}

#complete_image_set_list <- lapply(c(seq(1:nrow(flipped_count_table))), function(x) make_image_list(filtered_tree, flipped_count_table, test_tree_cut, x))

imd <- data.frame(sample_data(adults_rel_abund)) %>%
  rownames_to_column(var="sample_name")

image_metadata <- data.frame("sample_name" = rownames(flipped_count_table),
                             "image_list_index" = seq(1:nrow(flipped_count_table))) %>%
  left_join(imd)

saveRDS(image_metadata, "C:/Users/hammera/Documents/R_projects/Microbiopics/results/sample_pics/small_pic_test/microbe500_image_metadata.rds")
complete_image_set_list <- readRDS("C:/Users/hammera/Documents/R_projects/Microbiopics/results/sample_pics/small_pic_test/microbe500_image_list.rds")




```

```{r test_train_split}
set.seed(1975)
all_indices <- seq(1:nrow(flipped_count_table))
training_indices <- sample(all_indices, size=round(0.80*length(all_indices)), replace=F)
testing_indices <- all_indices[!(all_indices %in% training_indices)]

# need to generate a list of matrices for the data so that it's formatted like for other CNN problems

complete_image_set_list <- readRDS("C:/Users/hammera/Documents/R_projects/Microbiopics/results/sample_pics/small_pic_test/microbe500_image_list.rds")
View(as.matrix(complete_image_set_list))
str(complete_image_set_list)

```

```{r}
library(keras)
## really basic first-pass model architecture
model <- keras_model_sequential()
model %>%
 layer_conv_2d(filters = 32,
 kernel_size = c(3,3),
 activation = 'relu',
 input_shape = c(28,28,1)) %>%
 layer_conv_2d(filters = 64,
 kernel_size = c(3,3),
 activation = 'relu') %>%
 layer_max_pooling_2d(pool_size = c(2,2)) %>%
 layer_dropout(rate = 0.25) %>%
 layer_flatten() %>%
 layer_dense(units = 64, activation = 'relu') %>%
 layer_dropout(rate = 0.25) %>%
 layer_dense(units = 10, activation = 'softmax')

# note that if you're working with a 3d image you'll need to use a 3d convolutional layer
# Compile model
model %>% compile(loss = 'categorical_crossentropy',
 optimizer = optimizer_adadelta(),
 metrics = 'accuracy')

# Fit model
model_one <- model %>% fit(trainx,
 trainy,
 epochs = 15,
 batch_size = 128,
 validation_split = 0.2)
plot(model_one)

```








In the Pop-Phy-CNN paper they identify a subset of non-rare taxa to include in their analysis
The paper that they cite only takes the microbes that are in the top 500 in terms of abundance
This is one way to perform feature selection, but what if we took only the taxa that are useful for disease classification?

Obviously this would be scaled up for 



What are we bringing to the table that is novel?
1. Can we deal with zero inflation issues that result from taking a bunch of non-informative information?
  a. there is very likely an elbow point at which new information is not being assimilated
2. Can we build a unifying framework so that we get robust classification across disease states?
3. How does this model compare with other ML models that are trained on these data?
4. One other question that arises for me, has to do with the multi-omic embedding. 
5. Could we cluster metagenomic sequences in UniRef90 into protein families, then perform a convolutional operation on those for classification purposes?
  a. One idea, would be to first perform a 16s feature selection approach, then use the subtending metagenomic features for classification. Like what functions from disease-linked bugs are important for classification across diseases?
  b. a sub question that arises for me, is if there are consistent classes of functions that display robust association with diseases, which are from specific microbes that are linked to those diseases. For instance, if we find that there is specific mannose-metabolism genes from multiple bugs that are strongly linked to diabetes, is a quorum of functional association across microbes indicative of modules of interest?
6. One idea that I'm nearly certain would be novel, would be comparing the predictive import of various functional modules using CNNs or GNNs, where the data is structured according to the phylogenetic tree
  a. This would allow us to be more specific about the specific microbial functions that are associated with some disease outcome
  b. Additionally, it could clarify "how" microbes are doing what they're doing
  c. This same approach would have broad application across metagenome types
  d. do we have information that ties certain metagenomic modules to taxonomic mappings?



  








